---
title: 'SDS 322E Project 2 - Cleaning'
author: "Isabella Hsu"
date: "5/6/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Initializing Libraries
```{r}
library(tidyverse)
library(tibble)
library(dplyr)
library(ggplot2)
```

## Data Upload

```{r}
#uploading district data
district <- read.csv("DistrictType.csv")
#uploading staar data
staar <- read.csv("STAAR.csv")
```

## Data Cleaning

### Picking Relevant Variables

```{r}
#cleaning district data
district <- district %>%
  #removing descriptions - will just note these from district type codes
  select(-c(TEA.Description, NCES.Description)) %>%
  mutate(TEA.District.Type = as.factor(TEA.District.Type),
         NCES.District.Type = as.factor(NCES.District.Type)) %>%
  #renaming variables to be shorter
  rename(Name = District, District = District.Number, TEA = TEA.District.Type, NCES = NCES.District.Type, Charter = Charter.School..Y.N.)
```

```{r}
#cleaning STAAR data
staar <- staar %>%
  #removing apostrophes from district numbers
  separate(col = DISTRICT, into = c('apostrophe', 'District'), sep = "'") %>%
  select(-apostrophe) %>%
  #selecting specific data per population
  select(District, DDA03AMA1219R,	DDA03ARE1219R, DDE03AMA1219R, DDE03ARE1219R, DDF03AMA1219R,
         DDF03ARE1219R, DDM03AMA1219R,	DDM03ARE1219R,	DDR03AMA1219R,	DDR03ARE1219R,
         DDS03AMA1219R,	DDS03ARE1219R) %>%
  #renaming the variables to their descriptors
  rename(All.Math = DDA03AMA1219R,	
         All.Reading = DDA03ARE1219R,
         Econ_Dis.Math = DDE03AMA1219R,
         Econ_Dis.Reading = DDE03ARE1219R, 
         Female.Math = DDF03AMA1219R, 
         Female.Reading = DDF03ARE1219R, 
         Male.Math = DDM03AMA1219R,	
         Male.Reading = DDM03ARE1219R,	
         At_Risk.Math = DDR03AMA1219R,	
         At_Risk.Reading = DDR03ARE1219R,
         Special.Math = DDS03AMA1219R,	
         Special.Reading = DDS03ARE1219R)

glimpse(staar)

```

### Cleaning NAs 
```{r}
#Coding -1s and .s (which means NA) to NA
staar[staar == -1] <- NA
staar[staar == "."] <- NA

#Removing rows with na
staar <- staar %>%
  na.omit()

glimpse(staar)
```

### Joining the Data 

```{r}

#Making key a number type
staar <- staar %>%
  mutate(District = as.numeric(District))

#Joining two datasets together
staarjoin = left_join(staar, district, key = District)

#reordering 
staarjoin <- staarjoin %>%
  select(District, Name, TEA, NCES, Charter, All.Math, All.Reading, Econ_Dis.Math,   
         Econ_Dis.Reading,Female.Math, Female.Reading, Male.Math, Male.Reading, 
         At_Risk.Math, At_Risk.Reading, Special.Math, Special.Reading)

glimpse(staarjoin)
```


### Further Data Manipulation

```{r}
#pivoting the scores 
staar_pivot <- 
  staarjoin %>%
  pivot_longer(cols = c(All.Math, All.Reading, Econ_Dis.Math, Econ_Dis.Reading, 
                        Female.Math, Female.Reading, 
                         Male.Math, Male.Reading, At_Risk.Math, 
                        At_Risk.Reading,Special.Math, Special.Reading),
               names_to = c("Student_Population", "Subject"), 
               names_sep = "(\\.)",
               values_to = "Score")

```
*Please note that this pivot dataset is used to make the data tidy. However, for the purposes of further analysis, the previous joined dataset is better because it allows me to analyze academic outcomes for a specific segment without filtering.*

```{r}
#saving new datasets
write.csv(staarjoin, "staarjoin.csv")
write.csv(staar_pivot, "staar_pivot.csv")

```


